---
title: "Preliminary analysis 3 rust test 1"
subtitle: "The third step of prelim analysis"
author: "Firza Riany"
date: "8/21/2020"
output: 
  html_document:
    toc: TRUE
    number_sections: TRUE
---

# Description

In my experimental dataset, I had 154 poplar individuals infected by 3 pathogens that were spanning across 5 blocks. The intention of having the blocks is for repetition and to avoid bias. Because I wanted to use this dataset to model the association between genes and resistance components, I had to average the values (of the resistance components) across 5 blocks, to get only one value for each individual.

# Input

```{r input}
load(file = "prelim_step2.RData")  # RData prepared from prelim analysis step 2
```

# Libraries

```{r library}
library(ggplot2)  # To map the experimental design  
library(breedR)  # To model the source of phenotypic variation
```

# Observing block effects using mixed-effects model

I used mixed-effects model to analyze the effects of block and individuals to the variance of the resistance components. Mixed-effects model is a regular model used in animal and plant breeding. There are two parts in the mixed-effects model:

-   Fixed part: The factor/variable whose effect is directly affecting the variance of the response variable and whose effect that we would like to test the hypothesis with. In a simple term, I would say this is "a direct variable that affects our test"

-   Random part: The factors/variables whose effects are indirectly affecting the variance of response variable and whose effects that we would not consider to test our hypothesis. In a simple term, I would say this is "an indirect or background variable that can affect the variance of our response variable"

In this section, I wanted to adjust the values of my resistance components. In a simple word, I wanted to **eliminate block effects from the values of my resistance components**.

-   To do that, I used the resistance components as the response variables.

-   I fitted the resistance components to the fixed part of the model so that the model could find the average of the values of the resistance components in the form of an intercept.

-   I fitted the genotype and block as random effects because I considered them as the background effects that affect the variance in my resistance components.

------------------------------------------------------------------------

To check the significant effects of the individuals I did the following:

-   Made a null model where neither the individuals nor the blocks were fitted to the fixed part (the procedures above are the procedures to make a null model)

-   Made a model where individuals were fitted to the fixed part (genotype_fit)

-   Used chi-square test of the log likelihood between the null model and genotype_fit

-   If the test return a value below 0.05, it means that individuals have a significant effect on the variance in the resistance components

-   If the test return a value above 0.05, it means that individuals do not have a significant effect.

------------------------------------------------------------------------

To check the significant effects of the blocks, I will do the following:

-   A null model

-   A model where blocks were fitted to the fixed part (block_fit)

-   Used chi-square test of the log likelihood between the null model and block_fit

-   If the test return a value below 0.05, it means that blocks have a significant effect on the variance in the resistance components

-   If the test return a value above 0.05, it means that blocks do not have a significant effect.

```{r mixmod, message=FALSE, warning=FALSE}
# strain_list <- strain_list <- list(s09AX27_clean, s93JE3_clean, sP72_clean) 
# List: strain_list, 1 = 09AX27, 2 = 93JE3, 3 = P72
# strain_list is available in prelim analysis step 2, already saved as .RData

# I will run the model in a loop

# First, creating an empty list for each of the resistance components

LP_mix <- list()  # For latent period

UN_mix <- list() # For uredinia number

US_mix <- list() # For uredinia size

for (i in 1:length(strain_list)) {  # For i in 1:3 (because there are 3 pathogens)
  LP_mix[[i]] <- remlf90(fixed = Laten_24.02.15_gp ~ 1,  # Mixmod for latent period
                         random = ~ Genotype + Block,
                         data = strain_list[[i]], # For pathogen 1, then 2, then 3
                         method = "ai")
  UN_mix[[i]] <- remlf90(fixed = total_NbSS ~ 1,  # Mixmod for uredinia number
                         random = ~ Genotype + Block,
                         data = strain_list[[i]],
                         method = "ai")
  US_mix[[i]] <- remlf90(fixed = Taill_05.08.20_gp ~ 1, # Mixmod for uredinia size
                         random = ~ Genotype + Block,
                         data = strain_list[[i]],
                         method = "ai")
}

summary(LP_mix[[1]])  # LP mixed model for pathogen 09AX27 (coded 1 in the strain_list)
```

I used remlf90 from the breedR package.

How to read the results:

-   Intercept: the average latent period
-   Variance components: the result from random effects (in this case, individuals and blocks)
-   Estimated variance: the estimated variances associated to genotype and block. To calculate the contribution of genotype to the total variance in latent period:

```{r genotype to total var}
genot_var = LP_mix[[1]]$var[[1]] / (LP_mix[[1]]$var[[1]] + LP_mix[[1]]$var[[2]] +
                                      LP_mix[[1]]$var[[3]])

# Est. var genotype / (est. var genotype + block + residual)

genot_var
```

So, it means that individuals contribute to 27% of the variance in latent period. I was only checking LP_mix[[1]]. This list contains the estimated variance for poplar individuals *infected with pathogen 09AX27*. So, this result only applies to the individuals infected with 09AX27. FYI, LP_mix[[2]] contains the variance for the individuals infected with 93JE3, and LP_mix[[3]] with P72.

Then I wanted to see if individuals or blocks have significant effects. I will focus only on the resistance components latent period (LP) when the individuals are infected by pathogen 09AX27.

### Checking the effects of individuals

```{r fix = genotype}
genot_fit_LP <- remlf90(fixed = Laten_24.02.15_gp ~ Genotype,  # fitted = genotype
                        random =  ~ Block,
                        data = strain_list[[1]],  # dataset for pathogen 09AX27
                        method = "ai")

summary(genot_fit_LP)

# Function to check the significance of individuals and blocks

sig.fun = function(random, non_random, n_test){
  stat = 2*(random - non_random)
  
  sig = 1 - pchisq(stat, n_test)
  
  return(sig)
}

random_LP_09 = LP_mix[[1]]$fit$`-2logL` # Getting the log-likelihood from mixed model LP infected with pathogen 09AX27, genotype not in the fixed part

nonrand_LP_09 = genot_fit_LP$fit$`-2logL` # Getting the log-likelihood from mixed model LP infected with pathogen 09AX27, genotype in the fixed part

# Checking the significance of individuals
sig.fun(random = random_LP_09, non_random = nonrand_LP_09, n_test = 153)

```

Result: individuals have significant effects on the variance in latent period for individuals infected with pathogen 09AX27.

What about the block effects?

### Checking the effects of blocks

```{r fix = blocks}
block_fit_LP = remlf90(fixed = Laten_24.02.15_gp ~ Block, # fitted = blocks
                       random =  ~ Genotype,
                       data = strain_list[[1]], # LP for individuals infected with 09AX27
                       method = "ai")

summary(block_fit_LP)

# random_LP_09
nonrand2_LP_09 = block_fit_LP$fit$`-2logL` # Getting the log-likelihood from mixed model LP infected with pathogen 09AX27, block in the fixed part

# Checking the significance of blocks
sig.fun(random = random_LP_09, non_random = nonrand2_LP_09, n_test = 4)
```

Turned out blocks do not have a significant effects on the variance in latent period for the individuals infected with pathogen 09AX27.

This is a good indicator for experimental trial. It is expected that individuals should have significant effects on the variance of the resistance components. Moreover, the further analysis is to detect the extent of genetics effects to the resistance components. So, the result makes sense. Furthermore, an experimental design should not introduce a systematic effect to the resistance components. Systematic effects can be from blocks or sample arrangement in the experimental design. It introduces bias to the analysis. So, the ideal result from a good experimental design is non-significant effects from blocks, like how it is shown in the result in this analysis.

## Variogram of the residuals (Strain 09AX27 for all traits)

From LP_mix for strain 09AX27, residuals explain \~60% of phenotypic variance.\
Would be good to inspect the residuals using variogram.

### Latent Period

```{r variogram LP 09AX27}
variogram(LP_mix[[1]], coord = s09AX27_clean[, colnames(s09AX27_clean) %in% c("X", "Y")],
          R = 30)
```

Seem that residuals come from environmental variation (related to the coordinates where the genotypes were placed).\
Checking other models (UN and US).

### Uredinia Number

```{r UN_mix 09AX27}
summary(UN_mix[[1]])  # UN mixmod strain 09AX27

variogram(UN_mix[[1]], coord = s09AX27_clean[, colnames(s09AX27_clean) %in% c("X", "Y")], 
          R = 30)
```

### Uredinia Size

```{r US_mix 09AX27}
summary(US_mix[[1]])  # US mixmod strain 09AX27

variogram(US_mix[[1]], coord = s09AX27_clean[, colnames(s09AX27_clean) %in% c("X", "Y")], 
          R = 30)
```

## Variogram of the residuals (Strain 93JE3 for all traits)

### Latent Period

```{r LP_mix 93JE3}
summary(LP_mix[[2]])  # LP mixmod strain 93JE3

variogram(LP_mix[[2]], coord = s93JE3_clean[, colnames(s93JE3_clean) %in% c("X", "Y")], 
          R = 30)
```

### Uredinia Number

```{r UN_mix 93JE3}
summary(UN_mix[[2]])  # UN mixmod strain 93JE3

variogram(UN_mix[[2]], coord = s93JE3_clean[, colnames(s93JE3_clean) %in% c("X", "Y")], 
          R = 30)
```

### Uredinia size

```{r US_mix 93JE3}
summary(US_mix[[2]])  # US mixmod strain 93JE3

variogram(US_mix[[2]], coord = s93JE3_clean[, colnames(s93JE3_clean) %in% c("X", "Y")], 
          R = 30)
```

## Variogram of the residuals (Strain P72 for all traits)

### Latent Period

```{r LP_mix P72}
summary(LP_mix[[3]])  # LP mixmod P72

variogram(LP_mix[[3]], coord = sP72_clean[, colnames(sP72_clean) %in% c("X", "Y")], 
          R = 30)
```

### Uredinia Number

```{r UN_mix P72}
summary(UN_mix[[3]])  # UN mixmod strain P72

variogram(UN_mix[[3]], coord = sP72_clean[, colnames(sP72_clean) %in% c("X", "Y")], 
          R = 30)
```

### Uredinia Size

```{r US_mix P72}
summary(US_mix[[3]])  # US mixmod strain P72

variogram(US_mix[[3]], coord = sP72_clean[, colnames(sP72_clean) %in% c("X", "Y")], 
          R = 30)
```

# Adjusting the phenotypes from block effects

Because blocks are a systematic variation that affects the variation in phenotypes, we have to remove them.

```{r functions adj}
# Functions for adjustment ----
fun.coord <- function(X, Y){  # X and Y are coordinates from the dataset
  paste(X, Y, sep = "-")  # Using functions: X = dataset$X, Y = dataset$Y
}

# models: LP_mix (list of 3), UN_mix, US_mix, BP_mix
# LP_mix[[1]] = s09AX27, [[2]] = s93JE3, [[3]] sP72
intercept <- function(mix){  # mix being the name of model
    mix$fixed$Intercept[[1]][1, "value"]  # Example, mix = LP_mix[[1]]
}

BLUP.genot <- function(mix, x){  # x being the name of data frame object
  merge(x, data.frame("Genotype" = rownames(mix$ranef$Genotype[[1]]),
                      "BLUP_Genotype" = mix$ranef$Genotype[[1]]$value),
        by = "Genotype")
}

BLUP.block <- function(mix, x){
  merge(x, data.frame("Block" = rownames(mix$ranef$Block[[1]]),
                      "BLUP¨.block" = mix$ranef$Block[[1]]$value),
        by = "Block")
}


BLUP.resid <- function(data, clean, mix){  # clean and data being the name of data frame object
  merge(data, data.frame("coord" = fun.coord(clean$X, clean$Y),
                      "BLUP_resid" = residuals(mix)),
        by = "coord")
}

adj <- function(x){  # x being the name of data frame object
  x$intercept + x$BLUP_Genotype + x$BLUP_resid
}
```

## Adjusting LP

```{r adj LP}
adj_LP <- list(s09AX27_clean[, -c(6:15)],  # creating a list of dataframe to store loop result
               s93JE3_clean[, -c(6:15)],
               sP72_clean[, -c(6:15)])

for (x in 1:length(adj_LP)){  # x in data frame 1 - 3
  adj_LP[[x]]$coord <- fun.coord(X = adj_LP[[x]]$X, Y = adj_LP[[x]]$Y)  # adding coordinates to data frame x
  adj_LP[[x]]$intercept <- intercept(mix = LP_mix[[x]])  # adding intercept extracted from mixmodel
  adj_LP[[x]] <- BLUP.genot(mix = LP_mix[[x]], x = adj_LP[[x]])  # adding BLUP estimates extracted from mixmodel
  adj_LP[[x]] <- BLUP.block(mix = LP_mix[[x]], x= adj_LP[[x]])
  adj_LP[[x]] <- BLUP.resid(data = adj_LP[[x]], clean = adj_LP[[x]],  # adding residuals extracted from mixmodel
                             mix = LP_mix[[x]])
  adj_LP[[x]]$adj_LP <- adj(x = adj_LP[[x]])  # adjusting LP by removing block estimates
}
```

## Adjusting UN

```{r adj UN}
adj_UN <- list(s09AX27_clean[, -c(6:15)],
               s93JE3_clean[, -c(6:15)],
               sP72_clean[, -c(6:15)])

for (x in 1:length(adj_UN)){
  adj_UN[[x]]$coord <- fun.coord(adj_UN[[x]]$X, adj_UN[[x]]$Y)
  adj_UN[[x]]$intercept <- intercept(mix = UN_mix[[x]])
  adj_UN[[x]] <- BLUP.genot(UN_mix[[x]], adj_UN[[x]])
  adj_UN[[x]] <- BLUP.block(UN_mix[[x]], adj_UN[[x]])
  adj_UN[[x]] <- BLUP.resid(data = adj_UN[[x]], clean = adj_UN[[x]], mix = UN_mix[[x]])
  adj_UN[[x]]$adj_UN <- adj(adj_UN[[x]])
}
```

## Adjusting US

```{r adj US}
adj_US <- list(s09AX27_clean[, -c(6:15)],
               s93JE3_clean[, -c(6:15)],
               sP72_clean[, -c(6:15)])

for (x in 1:length(adj_US)){
  adj_US[[x]]$coord <- fun.coord(adj_US[[x]]$X, adj_US[[x]]$Y)
  adj_US[[x]]$intercept <- intercept(mix = US_mix[[x]])
  adj_US[[x]] <- BLUP.genot(US_mix[[x]], adj_US[[x]])
  adj_US[[x]] <- BLUP.block(US_mix[[x]], adj_US[[x]])
  adj_US[[x]] <- BLUP.resid(data = adj_US[[x]], clean = adj_US[[x]], mix = US_mix[[x]])
  adj_US[[x]]$adj_US <- adj(adj_US[[x]])
}
```

## Extracting the adjusted traits

```{r head 09AX27}
# Strain 09AX27
adj_s09AX27 <- s09AX27_clean[, -c(6:15)]
adj_s09AX27$Latence <- adj_LP[[1]]$adj_LP
adj_s09AX27$NbSS <- adj_UN[[1]]$adj_UN
adj_s09AX27$Taille <- adj_US[[1]]$adj_US

head(adj_s09AX27)  # preview of the dataset

par(mfrow = c(2,2))  
hist(adj_s09AX27$Latence)  # distribution of the adjusted traits
hist(adj_s09AX27$NbSS)
hist(adj_s09AX27$Taille)
```

```{r extract 09AX27, eval=FALSE, message=FALSE, warning=FALSE}
write.table(adj_s09AX27, "adj_traits_s09AX27.txt", sep = "\t", na = "NA", 
            quote = FALSE, row.names = FALSE)
```

```{r head 93JE3}
# Strain 93JE3
adj_s93JE3 <- s93JE3_clean[, -c(6:15)]
adj_s93JE3$Latence <- adj_LP[[2]]$adj_LP
adj_s93JE3$NbSS <- adj_UN[[2]]$adj_UN
adj_s93JE3$Taille <- adj_US[[2]]$adj_US

head(adj_s93JE3)

par(mfrow = c(2,2)) 
hist(adj_s93JE3$Latence)
hist(adj_s93JE3$NbSS)
hist(adj_s93JE3$Taille)
```

```{r extract 93JE3, message=FALSE, warning=FALSE, eval=FALSE}
write.table(adj_s93JE3, "adj_traits_s93JE3.txt", sep = "\t", na = "NA", 
            quote = FALSE, row.names = FALSE)
```

```{r head P72}
# Strain P72
adj_sP72 <- sP72_clean[, -c(6:15)]
adj_sP72$Latence <- adj_LP[[3]]$adj_LP
adj_sP72$NbSS <- adj_UN[[3]]$adj_UN
adj_sP72$Taille <- adj_US[[3]]$adj_US

head(adj_sP72)

par(mfrow = c(2,2)) 
hist(adj_sP72$Latence)
hist(adj_sP72$NbSS)
hist(adj_sP72$Taille)
```

```{r extract P72, eval=FALSE, message=FALSE, warning=FALSE}
write.table(adj_sP72, "adj_traits_sP72.txt", sep = "\t", na = "NA", 
            quote = FALSE, row.names = FALSE)
```

# Saving .RData for future analysis

```{r save RData, message=FALSE, warning=FALSE, eval=FALSE}
save(s09AX27_clean, s93JE3_clean, sP72_clean, strain_list,
     s09AX27_loop, s93JE3_loop, sP72_loop,
     adj_LP, adj_UN, adj_US,
     adj_s09AX27, adj_s93JE3, adj_sP72,
     file = "prelim_step3.RData")
```
